---
title: "K-M OS Plot Matching"
author: "David Dessert"
date: "Tuesday, February 25, 2015"
output: html_document
keep_md: true
---


Load the necessary packages to execute.  
[R Survival package](http://cran.r-project.org/web/packages/survival/survival.pdf)  
[ggplot2 documentation](http://docs.ggplot2.org/current/index.html)
```{r}
library(survival)       # package for survival plots
library(ggplot2)        # for plotting
library(grid)           # for sub-plots within plots
```


Define custom function to create a survival data.frame. This function is from [this](http://www.ceb-institute.org/bbs/wp-content/uploads/2011/09/handout_ggplot2.pdf) ggplot2 tutorial paper by Hadley Wickham of Rice University.
```{r}
createSurvivalFrame <- function(f.survfit)
{
    # initialise frame variable
    f.frame <- NULL
    # check if more then one strata
    if(length(names(f.survfit$strata)) == 0)
    {
        # create data.frame with data from survfit
        f.frame <- data.frame(time=f.survfit$time, n.risk=f.survfit$n.risk, n.event=f.survfit$n.event, n.censor = f.survfit
                              $n.censor, surv=f.survfit$surv, upper=f.survfit$upper, lower=f.survfit$lower)
        # create first two rows (start at 1)
        f.start <- data.frame(time=c(0, f.frame$time[1]), n.risk=c(f.survfit$n, f.survfit$n), n.event=c(0,0),
                              n.censor=c(0,0), surv=c(1,1), upper=c(1,1), lower=c(1,1))
        # add first row to dataset
        f.frame <- rbind(f.start, f.frame)
        # remove temporary data
        rm(f.start)
    }
    else 
    {
        # create vector for strata identification
        f.strata <- NULL
        for(f.i in 1:length(f.survfit$strata))
        {
            # add vector for one strata according to number of rows of strata
            f.strata <- c(f.strata, rep(names(f.survfit$strata)[f.i], f.survfit$strata[f.i]))
        }
        # create data.frame with data from survfit (create column for strata)
        f.frame <- data.frame(time=f.survfit$time, n.risk=f.survfit$n.risk, n.event=f.survfit$n.event, n.censor = f.survfit
                              $n.censor, surv=f.survfit$surv, upper=f.survfit$upper, lower=f.survfit$lower, strata=factor(f.strata))
        # remove temporary data
        rm(f.strata)
        # create first two rows (start at 1) for each strata
        for(f.i in 1:length(f.survfit$strata))
        {
            # take only subset for this strata from data
            f.subset <- subset(f.frame, strata==names(f.survfit$strata)[f.i])
            # create first two rows (time: 0, time of first event)
            f.start <- data.frame(time=c(0, f.subset$time[1]), n.risk=rep(f.survfit[f.i]$n, 2), n.event=c(0,0),
                                  n.censor=c(0,0), surv=c(1,1), upper=c(1,1), lower=c(1,1), strata=rep(names(f.survfit$strata)[f.i],
                                                                                                       2))
            # add first two rows to dataset
            f.frame <- rbind(f.start, f.frame)
            # remove temporary data
            rm(f.start, f.subset)
        }
        # reorder data
        f.frame <- f.frame[order(f.frame$strata, f.frame$time), ]
        
        
        # rename row.names
        rownames(f.frame) <- NULL
    }
    # return frame
    return(f.frame)
}
```


Define custom function to draw Kaplan-Meier curve with ggplot. This function is from [this](http://www.ceb-institute.org/bbs/wp-content/uploads/2011/09/handout_ggplot2.pdf) ggplot2 tutorial paper by Hadley Wickham of Rice University.
```{r}
qplot_survival <- function(f.frame, f.CI="default", f.shape=3)
{
    # use different plotting commands dependig whether or not strata's are given
    if("strata" %in% names(f.frame) == FALSE)
    {
        # confidence intervals are drawn if not specified otherwise
        if(f.CI=="default" | f.CI==TRUE )
        {
            # create plot with 4 layers (first 3 layers only events, last layer only censored)
            # hint: censoring data for multiple censoring events at timepoint are overplotted
            
            x <- as.vector(t(cbind(f.frame$time, 
                                          c(f.frame$time[-1], f.frame$time[NROW(f.frame$time)]))))
            y.upper <- as.vector(t(cbind(f.frame$upper, f.frame$upper)))
            y.lower <- as.vector(t(cbind(f.frame$lower, f.frame$lower)))
            ribbon <- data.frame(x, y.upper, y.lower)
            
            
            # (unlike in plot.survfit in survival package)
            ggplot(data=f.frame) + geom_step(aes(x=time, y=surv), direction="hv") + 
                geom_step(aes(x=time, y=upper), directions="hv", linetype=2) +   # upper 95% CI "---"
                geom_step(aes(x=time, y=lower), direction="hv", linetype=2) +    # lower 95% CI "---"
                geom_ribbon(data=ribbon, aes(x=x, ymin=y.lower, ymax=y.upper), alpha=0.15) +
                geom_point(data=subset(f.frame, n.censor==1), aes(x=time, y=surv), shape=f.shape) +  # censored events "+"
                ylim(0, 1)
        }
        else 
        {
            # create plot without confidence intervalls
            ggplot(data=f.frame) + geom_step(aes(x=time, y=surv), direction="hv") +
                geom_point(data=subset(f.frame, n.censor==1), aes(x=time, y=surv), shape=f.shape) +
                ylim(0, 1)
        }
    }
    else 
    {
        if(f.CI=="default" | f.CI==FALSE)
        {
            # without CI
            ggplot(data=f.frame, aes(group=strata, colour=strata)) + 
                geom_step(aes(x=time, y=surv), direction="hv") + 
                geom_point(data=subset(f.frame, n.censor==1), aes(x=time, y=surv), shape=f.shape) +
                ylim(0, 1)
        }
        else 
        {
            # with CI (hint: use alpha for CI)
            ggplot(data=f.frame, aes(colour=strata, group=strata)) + 
                geom_step(aes(x=time, y=surv), direction="hv") + 
                geom_step(aes(x=time, y=upper), directions="hv", linetype=2, alpha=0.5) +
                geom_step(aes(x=time,y=lower), direction="hv", linetype=2, alpha=0.5) +
                geom_point(data=subset(f.frame, n.censor==1), aes(x=time, y=surv), shape=f.shape) +
                ylim(0, 1)
        }
    }
}
```

Read in the .csv data file.

Table: CSV File contents

| Column   | Description     |
| ---------| ------------------- |
| months   | time of event  |
| OSpct    | OS percentage obtained from published K-M graph  |
| death    | 1 if death event  |
| Rcensor  | 1 if right-censored event  |
| n.risk   | at-risk number obtained from published K-M graph  |

Read in the K-M survival data extracted from the Figure 1A of the clinical trial results [paper](http://www.nejm.org/doi/pdf/10.1056/NEJMoa1011923)
```{r}
# paper.df <- read.csv('./NCT00112658/FOLFIRINOX_OS_data.csv')
paper.df <- read.csv('./NCT00112658/Gemcitabine_OS_data.csv')
head(paper.df)
```
Create a new data frame from the input CSV file for our work. The helper functions need 
data set up.

Table: Event definition

| Event   | Description        |
|:-------:| ------------------ |
| 0       | right-censored     |
| 1       | event (death)      |
| 2       | left-censored      |
| 3       | interval censored  |

```{r}
df <- data.frame(paper.df$months[is.na(paper.df$n.risk)], paper.df$death[is.na(paper.df$n.risk)])
names(df) <- c('months', 'event')

# Create a Survival object (package survival)
t.Surv <- Surv( time  = df$months, 
                event = df$event, 
                type  = 'right')          # interpret event as right-censored

# Create a Kaplan-Meier survival curve from the Survival object
t.survfit <- survfit(formula = t.Surv ~ 1, data = df)

# create frame from survival data (helper function defined above)
# t.survframe <- createSurvivalFrame(t.survfit)

# create kaplan-meier plot with ggplot (helper function defined above)
# qplot_survival(t.survframe)
```


Let's evaluate and tweak the fit to the K-M plot from the paper (only at the times where there were any events).
```{r}
# Evaluate at the paper's time intervals
paper.x <- c(0, paper.df$months[is.na(paper.df$n.risk)])
paper <- approx(x=c(0, paper.df$months[!is.na(paper.df$OSpct)]),
                y=c(1, paper.df$OSpct[!is.na(paper.df$OSpct)]/100),
                xout=paper.x,
                method='constant',
                rule=2)

# Interpolate the generated fit survival curve (continuous step)
# Note that I've added an initial point at time 0, 100% survival
fit <- approx(x=c(0, t.survfit$time), 
              y=c(1, t.survfit$surv), 
              xout=paper.x, 
              method='constant',
              rule=2)

# Calculating an MSE to track the error
err.fit <- paper$y - fit$y
mse <- mean(err.fit^2)
```

The MSE is `r mse`.  
Find all times such that $\deltaerr < 1 / $ `r -2*t.survfit$n` 
```{r}
err.fit.dy <- diff(err.fit)

# Find the times where err.fit.dy < -1/(2* #patients)
event.add <- err.fit.dy < -1/(2 * t.survfit$n)
```

Plot the paper's and the fitted K-M plots. Below, show the error between them.
```{r}
# The extra x, y points are to implement a step function
# for paper$x, paper$y
df_plot <- data.frame(as.vector(t(cbind(paper$x, c(paper$x[-1], paper$x[NROW(paper$x)])))), 
                      as.vector(t(cbind(paper$y, paper$y))), 
                      factor(x='paper', levels=c('paper', 'fit')))
names(df_plot) <- c('x','y', 'type')

# Extra points for step plot of fit$x, fit$y
df_plot <- rbind(df_plot, 
                 data.frame(x=as.vector(t(cbind(fit$x, c(fit$x[-1], paper$x[NROW(fit$x)])))), 
                            y=as.vector(t(cbind(fit$y, fit$y))), 
                            type='fit'))

grid.newpage()
pushViewport(viewport(layout=grid.layout(4, 1)))

# Upper subplot
p.a <- ggplot(df_plot, aes(x=x, y=y, colour=type)) + geom_line()
for (idx in which(event.add))
{
    p.a <- p.a + annotate("point", 
                          x=paper$x[idx+1], y=paper$y[idx+1], 
                          colour='black', size=5, alpha=0.25)
}
p.a <- p.a + annotate("text", x=max(paper$x), y=0.9, label=sprintf('MSE=%e', mse[NROW(mse)]), hjust=1, vjust=1)
p.a <- p.a + xlab('months')
p.a <- p.a + ylab('Probability (%)')
p.a <- p.a + ggtitle('K-M Plot Fit (Paper vs Fit)')

# Lower subplot
p.b <- qplot(paper$x, paper$y-fit$y)
p.b <- p.b + annotate("segment", x=0, xend=max(paper$x), y=sqrt(mse[NROW(mse)]), yend=sqrt(mse[NROW(mse)]), color='red')
p.b <- p.b + annotate("segment", x=0, xend=max(paper$x), y=-sqrt(mse[NROW(mse)]), yend=-sqrt(mse[NROW(mse)]), color='red')
for (idx in which(event.add))
{
    p.b <- p.b + annotate("segment", 
                          x=paper$x[idx], xend=paper$x[idx+1], 
                          y=err.fit[idx], yend=err.fit[idx+1], 
                          colour='red')
}
p.b <- p.b + xlab('months') + ylab('Error')

print(p.a, vp=viewport(layout.pos.row=1:3, layout.pos.col=1))
print(p.b, vp=viewport(layout.pos.row=4, layout.pos.col=1))
```

Add death events at these indices
```{r}
df <- rbind(df, data.frame(months=paper$x[which(event.add)+1], event=1))

# Redo the analysis
t.Surv <- Surv( time  = df$months, 
                event = df$event, 
                type  = 'right')          # interpret event as right-censored

t.survfit <- survfit(formula = t.Surv ~ 1, data = df)

# create frame from survival data
# t.survframe <- createSurvivalFrame(t.survfit)

# create kaplan-meier plot with ggplot
# qplot_survival(t.survframe)

# Let's re-evaluate the updated fit to the K-M plot in the paper
paper.x <- c(0, paper.df$months[is.na(paper.df$n.risk)])
paper <- approx(x=c(0, paper.df$months[!is.na(paper.df$OSpct)]),
                y=c(1, paper.df$OSpct[!is.na(paper.df$OSpct)]/100),
                xout=paper.x,
                method='constant',
                rule=2)

# Interpolate the generated fit survival curve (continuous step)
# Note that I've added an initial point at time 0, 100% survival
fit <- approx(x=c(0, t.survfit$time), 
              y=c(1, t.survfit$surv), 
              xout=paper.x, 
              method='constant',
              rule=2)

# Calculating an MSE to track the error
err.fit <- paper$y - fit$y
mse <- mean(err.fit^2)
```

The MSE is `r mse`.  
Find all times such that $\deltaerr < 1 / $ `r -2*t.survfit$n` 
```{r}
err.fit.dy <- diff(err.fit)

# Find the times where err.fit.dy < -1/(2* #patients)
event.add <- err.fit.dy < -1/(2 * t.survfit$n)
```

```{r}   
df_plot <- data.frame(paper$x, paper$y, factor(x='paper', levels=c('paper', 'fit')))
names(df_plot) <- c('x','y', 'type')

df_plot <- rbind(df_plot,
                 data.frame(x=fit$x, y=fit$y, type='fit'))

grid.newpage()
pushViewport(viewport(layout=grid.layout(4, 1)))

# plot the difference between the two plots
p.a <- ggplot(df_plot, aes(x=x, y=y, colour=type)) + geom_step(direction='hv')
for (idx in which(event.add))
{
    p.a <- p.a + annotate("point", 
                          x=paper$x[idx], y=paper$y[idx], 
                          colour='black', size=5, alpha=0.25)
}
p.a <- p.a + annotate("text", x=max(paper$x), y=0.9, label=sprintf('MSE=%e', mse[NROW(mse)]), hjust=1, vjust=1)
p.a <- p.a + xlab('months')
p.a <- p.a + ylab('Probability (%)')
p.a <- p.a + ggtitle('K-M Plot Fit (Paper vs Fit)')

# Lower subplot
p.b <- qplot(paper$x, paper$y-fit$y)
p.b <- p.b + annotate("segment", x=0, xend=max(paper$x), y=sqrt(mse[NROW(mse)]), yend=sqrt(mse[NROW(mse)]), color='red')
p.b <- p.b + annotate("segment", x=0, xend=max(paper$x), y=-sqrt(mse[NROW(mse)]), yend=-sqrt(mse[NROW(mse)]), color='red')
for (idx in which(event.add))
{
    p.b <- p.b + annotate("segment", 
                          x=paper$x[idx], xend=paper$x[idx+1], 
                          y=err.fit[idx], yend=err.fit[idx+1], 
                          colour='red')
}
p.b <- p.b + xlab('months') + ylab('Error')

print(p.a, vp=viewport(layout.pos.row=1:3, layout.pos.col=1))
print(p.b, vp=viewport(layout.pos.row=4, layout.pos.col=1))
```

Manually add the right-censored events. I don't have an automatic way to do this yet. I can use the paper.df$n.risk numbers to know which time regions need a right-censored event added. Also, because this plot showed all the right-censored events (black tick-marks), I would only add right-censored events where previous ones had been recorded.
```{r}
# Right-censored events added to df
df <- rbind(df, data.frame(months=c(1.98, 5.63, 6.15), event=0))
```

Regenerate the K-M plot from the paper.
```{r}
# Redo the survival analysis (final time)
t.Surv <- Surv( time  = df$months, 
                event = df$event, 
                type  = 'right')          # interpret event as right-censored

t.survfit <- survfit(formula = t.Surv ~ 1, data = df)

t.survframe <- createSurvivalFrame(t.survfit)

# K-M Survival plot
qplot_survival(t.survframe)
p <- last_plot() + xlab("Months") + ylab("Probablility") + ggtitle("Overall Survival")
p <- p + annotate("text", x=20, y=c(1.00, 0.95, 0.90, 0.85, 0.80), hjust=0, vjust=1,
                  label=c("FOLFIRINOX",
                          "Metastatic Patients",
                          "First-Line Treatment",
                          "NCT00112658",
                          "http://www.nejm.org/doi/pdf/10.1056/NEJMoa1011923"))
print(p)

# Histogram plot
p <- ggplot(df, aes(x=months)) + geom_histogram(aes(y= ..density..), binwidth=2) + geom_density()
p <- p + xlab("Months") + ylab("Deaths") + ggtitle("Overall Survival")
p <- p + annotate("text", x=20, y=c(0.070, 0.066, 0.062, 0.058, 0.054), hjust=0, vjust=1,
                  label=c("FOLFIRINOX",
                          "Metastatic Patients",
                          "First-Line Treatment",
                          "NCT00112658",
                          "http://www.nejm.org/doi/pdf/10.1056/NEJMoa1011923"))
print(p)


r <- hist(t.survframe$time, breaks=42, plot=FALSE)
r$density[41] <- sum(r$density[41:NROW(r$density)])
r$density <- r$density[-NROW(r$density):-42]
r$counts[41] <- sum(r$counts[41:NROW(r$counts)])
r$counts <- r$counts[-NROW(r$counts):-42]
r$breaks <- 0:41
r$mids <- 0.5:40.5
plot(r, freq=FALSE, xlab="Months", main="FOLFIRINOX OS", sub="First-Line Treatment", ylim=c(0, 0.1))


```
